Wormtable: A data structure for genome scale data

### 2013-07-13 docs/paper/paper.tex
Good draft.

\begin{abstract}
\section{Summary:}
Biological research generates vast quantities of tabular data, which must 
be processed to test hypotheses. These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is an extremely inefficient process.
Furthermore, there is no simple means of indexing these files so that 
arbitrary data values within rows can be quickly found. We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables.
\section{Availability:}
Package available at 
\href{http://pypi.python.org/pypi/wormtable}{http://pypi.python.org/pypi/wormtable}

@@

\section{Introduction}
% Lots of data
The improvement in sequencing technologies has
lead to an exponential increase in the volume of data being managed 
by biologists. Standard studies require the storage and processing 
of hundreds of gigabytes of data, and this volume of data is only set to 
increase as sequencing technologies continue to improve.

% but we use text files
Despite the increasing importance of data processing, however, the methods 
used in Bioinformatics are almost entirely 
based around the use of text files. Plain text files have many advantages:
they are extremely portable, easy to view and interpret and can be processed on 
any platform with the minimum of library dependencies. Text files also 
have many disadvantages however; the principle difficulties for our 
purposes here are, they are slow to parse and difficult to index.

% text files are also easy to read/comprehend and don't require specialist
% knowledge to process.
% I think learning a database syntax is a barrier to many researchers that are
% not trained bioinformaticians

% Databases solve these problems
The classical approach to working with large volumes of 
structured data is to use a database, which can solve these 
problems very effectively.
Data is typically encoded in an efficient binary 
format, saving substantial space and avoiding the need
for repeated parsing. Records are usually stored in a
B-tree, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features.

% there are a range of standard (well defined) tabular data formats: vcf, gtf
% lends itself to a database solution geared towards specific file types

% Databases but introduce several others. 
One may, of course, load textual data into a relational database, and 
use SQL to retrieve the records of interest. This approach has several 
disadvantages, however. Firstly, one must first examine the data 
and construct an appropriate SQL schema to define the appropriate columns. 
It is sometimes feasible to do this automatically 
for data files, but is certainly not a trivial process, 
particularly if the schema 
must be portable across different relational database platforms.
Even assuming that the data 
has been inserted into a table with an appropriate schema, there 
are still difficulties to be addressed. Maintaining a database 
server is a difficult task and when the volumes of data become
large, a specialised administrator is usually required to 
ensure the server works correctly.

% We don't need concurrency, etc, as it's always read-only.
Much of the time, a database server is far more complicated than
we require for biological data. Data files 
are usually written once by an application 
and do not subsequently change. Thus, storing this information 
in a relational database with its sophisticated concurrency control
and guarantees of consistency in the face of asynchronous updates
is entirely unnecessary. Interacting with a database is also a
substantial hurdle, as another lanuage must be learned in order 
to gain access to data. Relational databases, although providing 
powerful data management technologies are far more complex than 
is required, and are therefore little used for the types of 
data that we are interested in.

% are other database solutions less flexible when it comes to where the 
% data is stored?

\section{Wormtable}
% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. Tables can be read 
concurrently by any number of processes, and are accessed by a 
simple API in Python.

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of a relational database 
server, wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store~\citep{m12}. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. DB is a highly modular library, providing the  % do you mean Berkeley DB?
means to define a very lightweight read-only database, without the 
overheads of a server process

% Wormtable's binary row format 
On this solid foundation of row storage in Berkeley DB, wormtable uses a
compact binary row format designed specifically for biological applications.
Each row contains data from a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Supported types are signed and unsigned integers, 
floating point and character data. To ensure that values are 
stored as compactly as possible, wormtable supports integer values from 
$1$ to $8$ bytes, and the standard single and double precision 
floating point values.

% Wormtable's indexes 
An essential element of database technology is the ability to index tables
to rows with certain properties to be quickly retrieved. Wormtable 
supports indexing over columns or groups of columns within a table, 
making many operations extremely efficient.

\section{Examples}

% What is VCF?
A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
or VCF. In a VCF file, information about variant sites in a genome is encoded 
as tab-delimited rows in a text file. To access information about a 
particular site naively, one must proceed through the file line-by-line until the 
desired record is found. There are methods available to index this 
file~\citep{li11} by genomic position, but these have no semantic 
understanding of the structure of the file and so are limited in the 
fields that can be indexed. One cannot, for example, build an index 
to quickly obtain all records in which the quality field is greater
than some value. Such queries can take a considerable time on large VCF 
files, which discourages testing and checking for the effects of 
arbitrary thresholds in analyses.

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples of wormtable
that illustrate the improvements over existing methods. We applied these 
examples to a large, publicly available VCF file produced as part of the
Drosophila genetic reference panel~\citep{mr+12}.
This VCF\footnote{\href%
{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
contains variant calls 
from a population consisting 192 inbred Drosophila lines and contains data 
from the whole genome over a total of 6,146,611 rows and consists
of 15GB of uncompressed text. The VCF was converted to 
wormtable format using the included \texttt{vcf2wt} program
on a standard workstation with a single hard disk. The
conversion required approximately 70 minutes, and the size of the resulting  
wormtable file was 10.1GB (using the smallest type required
to represent the data in each column).

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable, standard Unix tools and PyVCF\footnote{%
\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
a Python VCF parser.
In each examples we proceed 
row-by-row, checking the REF and ALT columns for the type of variant and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 130 minutes. The Unix tools 
(\texttt{cut}, \texttt{grep} and \texttt{perl}) method required 
1m17s, and wormtable required 55 seconds. Both the Unix tools 
and wormtable were limited by the speed of the hard disk in the
test machine, and therefore could be much faster using 
modern solid state 
storage. It should be noted, however, that this simple task represents 
something of a best case scenario for the Unix tools pipeline.

It is well known, of course, that plain text cannot be retrieved 
efficiently at large scales. The BVCF format is a binary version 
of VCF which is intended to avoid this overhead [citation?]. 
This binary format does alleviate the problem of parsing overhead,
but does not solve the problem of indexing \emph{within} rows.

To illustrate the use of indexes in wormtable, we created an index 
on the REF and ALT columns which required 3m42s to build and 
consumed 66MB of storage space. Using this index we repeated the 
example of counting transitions and transversions, which required 
less than a second to complete.

The examples in this section are intended to illustrate the 
magnitude of performance gains that wormtable can provide 
when working with large datasets. The gains are many: we 
do not have to parse the information before it is used;
we can quickly seek to arbitrary rows within a table; and 
we can use indexes to intelligently sort and aggregate our
data to make it more useful.

\section{Conclusions}
% text files are good for archiving, but terrible for processing
Text files are the oldest and most reliable means of exchanging 
information between computers. Plain text, however, does not provide an efficient 
means of storing and processing large volumes of data. Current practice 
in bioinformatics is to encode data in more and more complex plain 
text data formats. While these formats are excellent from the perspective 
of minimising software dependencies, they are very inefficient in terms of 
storage space, the time required to parse them, and the linear
seek times to find particular rows.

% We're not trying to define a universal bioinformatics file format,
% just to help people be more productive.
Wormtable is not intended to replace text files as the universal 
interchange format. It is intended to provide a persistent data structure 
that efficiently holds data in a form that can be efficiently processed 
and searched. Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write simple code that processes this 
data in a very high performance manner. 
By providing conversion programs to convert 
common formats to wormtable, we also provide an easy to 
use common platform for 
bioinformatics analysis. This reduces the need for third party libraries 
to parse complex files and considerably simplifies the code required 
to process data.
Together, these advantages of increased performance and reduced 
code complexity can substantially increase a researcher's 
productivity and ability to explore their data.

>>>>

\begin{abstract}
\section{Summary:}
Biological research generates large amounts of tabular data.
These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is a time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables. 
\section{Availability:}
Package available at 
\href{http://pypi.python.org/pypi/wormtable}{http://pypi.python.org/pypi/wormtable}

@@

\section{Introduction}
Despite the ever increasing volumes of data being processed in bioinformatics,
the methods used are almost entirely based on plain text files. Data is 
usually encoded in rows, with each row consisting of a series of tab-delimited 
values.
These files are easy to view and interpret and can be processed on 
any platform with the minimum of library dependencies. Using compression,
text files can be quite compact, and Tabix~\citep{li11} provides  
fast retrieval of rows from arbitrary locations in the table.

It is not sufficient, however, to simply store and retrieve data. To use data
it must be \emph{processed}. This is the major flaw in 
using text files as a data format: before we can perform calculations, we 
must first parse the encoded information into a native 
machine value. Compression adds another substantial overhead. The result 
of this is that simple calculations over a large dataset may take 
many hours to complete.

Another problem with tabular text files is that it is difficult 
to index the information within partiular columns. 
This means that most operations on the table require a complete scan 
through the file. As a result, the only viable means of 
working with a subset of the data
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

The obvious solution to these problems is to load tabular data into a
relational database. Databases store values in binary form so that parsing is 
not required, and support efficient retrieval and indexing. There 
are many problems with this approach, however. Relational databases are 
extremely complex systems, each supporting different features and SQL 
dialects. It is not a straightforward task to design a schema for a 
particular data set, particularly not if portability accross different 
databases is required. Similiarly, accessing the data requires a knowledge
of SQL. In the most common case, a database server must also be maintained
and so storage and user permissons must be carefully managed. All of 
these aspects require significant expertise.

A relational database server, however, is far more than we require in the 
majority of cases. Data files are usually written once and 
not subsequently expected to change.
Thus, storing this information 
in a relational database with its sophisticated concurrency control
is entirely unnecessary. Centralised storage of data sets
creates an unnecessary administration burden, as does an extra
layer of user management. Therefore, although relational databases
provide powerful data management technologies, they are complex
to use and maintain and are therefore rarely used for 
static datasets.

\section{Wormtable}
% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data.

% Wormtable's binary row format 
Data is stored using a  
compact binary format designed specifically for biological applications.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Wormtable currently supports signed and unsigned integers, 
floating point and character data. To ensure that values are 
stored as compactly as possible, integer sizes can range from  
$1$ to $8$ bytes. The standard single and double precision 
floating point values are also supported, along with fixed and 
variable length strings.

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of relational databases,
wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is an open source embedded database toolkit that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. Berkeley DB is a highly modular library, providing the 
means to define a very lightweight read-only database, without the 
overheads and complexity of a relational database system.

Combining wormtable's custom row format with Berkeley DB gives us a 
powerful platform for data storage and processing.
An important advantage
of this combination is the ability to index data. Wormtable supports
indexes over arbitary combinations of columns, along with a flexibile
system of `binning' of values in an index.
Indexes can often be used to avoid the cost of a full table scan, and 
can therefore make many operations much faster.

\section{Examples}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\citep{da+11}.  
To illustrate the advantages of wormtable, we
converted a large 
publicly available VCF file, produced as part of the
Drosophila genetic reference panel~\citep{mr+12}, to wormtable format.
This VCF\footnote{\href%
{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a standard workstation with a single hard disk. The
conversion required approximately 68 minutes, and the size of the resulting  
wormtable file was 10.1GB (using the smallest type required
to represent the data in each column).

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF\footnote{%
\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
a Python VCF parser.
In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 130 minutes;
the wormtable version required
55 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 77 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that this simple task represents 
something of a best case scenario for the Unix tools pipeline.

Another major advantage of wormtable is the ability to create 
an index over arbitary combinations of columns within a table.
Indexes make many operations much more efficient, as they provide 
a way to quickly find particular rows and avoid the 
heavy cost of a full table scan.
To illustrate this advantage, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m42s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

\section{Conclusions}
It is well known that parsing text files is a very inefficient process, 
and binary file formats are preferable for this reason. The BCF format,
for example, is a binary version of the VCF format discussed above which 
is designed to avoid just this problem. There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for several different file formats is surely 
unsustainable.

Wormtable alleviates the need for these different formats and libraries,
as it is flexible enough to store many different types of data. Python 
is currently the only programming language supported, but a C library
and bindings for other languages are planned. Wormtable is very portable,
and has been tested on big- and little-endian platforms with 32 and 64
bit word sizes, along with many operating system combinations. To 
take advantage of the advanced data processing features of wormtable
all that is required is a conversion program, a considerably simpler
task than designing and supporting a custom binary file format.

Wormtable is not intended to replace text files as the universal 
interchange format for biological data. 
It is intended to provide a persistent data structure 
that can be efficiently processed 
and searched. Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write simple code that processes this 
data in a very high performance manner. 
Different file formats can be handled consistently 
in wormtable, reducing the need for third party libraries 
to parse complex files and simplifying the code required 
to process data.
Together, these advantages of increased performance and reduced 
code complexity can substantially increase a researcher's 
productivity and ability to explore their data.
